One-Particle Density Matrix Polarisation Susceptibility Tensors

### 2018-08-09 papers/1_dmatpol/submit/paper-submit.tex
Final review I.

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

@@

Molecular\hyp{}level exploration of condensed phase phenomena becomes increasingly unavoidable
as technology and medicine continue to develop. Unfortunately, the computational cost of accurate
electronic structure calculations grows extremely fast with the number of particles, greatly limiting
the use of the state\hyp{}of\hyp{}the\hyp{}art first\hyp{}principles theories when large and complex 
chemical systems such as solutions and biomolecules come into play.\cite{
Xu.Guidez.Bertoni.Gordon.JCP.2018,Tomasi.Mennucci.Cammi.ChemRev.2005}
Due to this reason, it has been of particular importance nowadays to develop bridging theoretical techniques 
that link extremely accurate but localised pieces of information with certain knowledge of an atomistic 
organisation and intermolecular interaction on a relatively large spatial scales, typically ranging beyond 
few nanometers.

The development of hybrid methods such as quantum mechanical\hyp{}molecular mechanical (QM/MM) technique 
was a revolution in computational chemistry which allows insightful theoretical studies of quantum mechanical
effects in biological systems because costly quantum mechanical calculations can be combined with lower\hyp{}level 
and much less expensive classical dynamics.\cite{Warshel.Levitt.JMolBiol.1976,Senn.Thiel.Angew.2009}
Fruitful incorporation of accurate \emph{ab initio} methods with extensive experimental and semi\hyp{}empirical
refinings led to development of a wide variety of force fields,\cite{Demerdash.Yap.T-Head-Gordon.AnnuRevPhysChem.2014}
whereas sophisticated partitioning of Hamiltonian or electron density enabled numerous schemes for efficient fragmentation
within a QM level.\cite{Gordon.Fedorov.Pruitt.Slipchenko.ChemRev.2012}
%Gao.Truhlar.Wang.Mazack.Loffler.Provorse.Rehak.ACR.2014}

@@

Now, if the AO basis is orthogonal, Eq.~\eqref{e:dmu-4-exact.linear-approximation} can be simplified by realising that
%
\begin{equation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{M}} \cdot \delta {\bf C}
   \right] }  
-
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{K}} \cdot \delta {\bf C}
   \right] } \;.
\end{equation}
%
Here we introduced the auxiliary tensors
%
\begin{subequations}
 \begin{align}
   \widetilde{\mathbb{M}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}}     \;,           \\
   \widetilde{\mathbb{K}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}} \cdot {\bf D}^{(0)} \;, \\
   \widetilde{\mathbb{L}}  &\equiv \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \;
 \end{align}
\end{subequations}
%
to simplify the notation. Hence, we have
%
\begin{equation} \label{e:dmu-l-vector}
  \frac{1}{4} 
 \delta {\BM{\upmu}} %({\bf r}_0) 
   =
   \Tr{ 
    \left[ 
         \widetilde{\mathbb{L}} \cdot \delta {\bf C}
    \right] }
   %
   \equiv \sum_{i}^{\rm occ} \sum_\alpha {\bf l}_{i\alpha} \delta C_{\alpha i} \;,
\end{equation}
%
where the Cartesian vector
%
\begin{equation}\label{e:l-vector}
 {\bf l}_{i\alpha} \equiv \left[ \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \right]_{i\alpha} 
      = \left[  {\bf C}^\dagger \cdot \mathbb{M} \cdot \left( {\bf 1} - {\bf D}^{(0)} \right) \right]_{i\alpha}
\end{equation}
%
and $i$ runs over occupied unperturbed MO's.
Note that the intermediate quantity $\widetilde{\mathbb{L}}$ contains the projector onto
the unoccupied MO space at the absence of the external electric field. It means that the
perturbed solution is constructed from the unperturbed virtual orbitals as well.
Although Eq.~\eqref{e:dmu-l-vector}
is exact up to first order in ${\BM\Delta}$, it cannot be inverted at this moment to obtain $\delta {\bf C}$
due to the presence of trace operation, unless the electric field is uniform.
Let the induced dipole moment be partitioned into separate contributions
from the occupied localised MO's (LMO's), i.e.,
%
\begin{equation} \label{e:dmu-part}
 \delta {\BM{\upmu}} = \sum_i^{\rm occ} \delta {\BM{\upmu}}_i \;,
\end{equation}
%
where
%
\begin{equation} \label{e:dmu-dpol}
 \delta {\BM{\upmu}}_i = {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) \;.
\end{equation}
%
In Eq.~\eqref{e:dmu-dpol}, ${\BM{\alpha}}_i$ is the distributed dipole\hyp{}dipole polarisability tensor
associated with the $i$th LMO whereas ${\bf F}({\bf r}_i)$ 
is the electric field evaluated at the $i$th LMO charge centroid,
${\bf r}_i = \tbraket{\phi_i}{\hat{\bf r}}{\phi_i}$.
By inserting the above result into Eq.~\eqref{e:dmu-l-vector},
recasting the right hand side in a matrix form
and assuming real quantities,
one can separate the contributions coming from occupied MO's as
%
\begin{equation} \label{e:dmu-l-vector-mo-transform-X}
 \frac{1}{4} {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) 
   =
   \delta {\bf c}_i^T \cdot {\bf L}_i \;.
\end{equation}
%
Note that now ${\bf L}_i$ is a matrix of dimension $n \times 3$ and $\delta{\bf c}_i$ is the $i$th column of the
change in the LCAO-MO matrix due to the polarisation process.
The solution for $\delta {\bf C}$ is
%
\begin{equation} \label{e:mu-ind.matrix.linear.solution}
  \delta {\bf c}_i^T = \frac{1}{4}
            {\bf F}({\bf r}_i)  \cdot {\BM{\alpha}}_i \cdot 
                    \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \;,
\end{equation}
%
where $\left[ {\bf L}_i  \right]^{-1}_{\rm Left}$ is a left inverse
of ${\bf L}_i$ matrix,
%
\begin{equation} 
      \left[ {\bf L}_i  \right]^{-1}_{\rm Left}   \equiv
       \left[ {\bf L}_i^T \cdot {\bf L}_i \right]^{-1} \cdot {\bf L}_i^T   
\end{equation}
%
with the dimension of $3\times n$. Note that the right inverse of ${\bf L}_i$
does not exist because ${\bf L}_i \cdot {\bf L}_i^T$ is singular due to the fact that usually
$n > 3$.

@@

\begin{acknowledgments}
This project is carried out under POLONEZ programme which has received funding from the European Union's
Horizon~2020 research and innovation programme under the Marie Skłodowska-Curie grant agreement 
No.~665778. This project is funded by National Science Centre, Poland 
(grant~no. 2016/23/P/ST4/01720) within the POLONEZ 3 fellowship.
%Many thanks to Robert W.~G{\'o}ra and Robert Zale{\'s}ny 
%for reading and commenting on the manuscript.
\end{acknowledgments}

>>>>

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage{txfonts}
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

% EU logo
\newcommand*{\InlineImage}[1]{%
    \raisebox{-.02\baselineskip}{%
        \includegraphics[
        height=\baselineskip,
        width=\baselineskip,
        keepaspectratio,
        ]{#1}%
    }%
}

@@

Molecular\hyp{}level exploration of condensed phase phenomena becomes increasingly unavoidable
as technology and medicine continue to develop. Unfortunately, the computational cost of accurate
electronic structure calculations grows extremely fast with the number of particles, greatly limiting
the use of the state\hyp{}of\hyp{}the\hyp{}art first\hyp{}principles theories when large and complex 
chemical systems such as solutions and biomolecules come into play.\cite{
Xu.Guidez.Bertoni.Gordon.JCP.2018,Tomasi.Mennucci.Cammi.ChemRev.2005}
Due to this reason, it has been of particular importance nowadays to develop bridging theoretical techniques 
that link extremely accurate but localised pieces of information with certain knowledge of an atomistic 
organisation and intermolecular interaction on relatively large spatial scales, typically ranging beyond 
few nanometers.

The development of hybrid methods such as quantum mechanical\hyp{}molecular mechanical (QM/MM) technique 
was a revolution in computational chemistry which allows insightful theoretical studies of quantum mechanical
effects in biological systems because costly quantum mechanical calculations can be combined with lower\hyp{}level 
and much less expensive classical dynamics.\cite{Warshel.Levitt.JMolBiol.1976,Senn.Thiel.Angew.2009}
Combination of accurate \emph{ab initio} methods with extensive experimental and semi\hyp{}empirical
refinings led to development of a wide variety of force fields,\cite{Demerdash.Yap.T-Head-Gordon.AnnuRevPhysChem.2014}
whereas sophisticated partitioning of Hamiltonian or electron density enabled numerous schemes for efficient fragmentation
within a QM level.\cite{Gordon.Fedorov.Pruitt.Slipchenko.ChemRev.2012}
%Gao.Truhlar.Wang.Mazack.Loffler.Provorse.Rehak.ACR.2014}

@@

Now, if the AO basis is orthogonal, Eq.~\eqref{e:dmu-4-exact.linear-approximation} can be simplified by realising that
%
\begin{equation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{M}} \cdot \delta {\bf C}
   \right] }  
-
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{K}} \cdot \delta {\bf C}
   \right] } \;.
\end{equation}
%
Here we introduced the auxiliary tensors
%
\begin{subequations}
 \begin{align}
   \widetilde{\mathbb{M}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}}     \;,           \\
   \widetilde{\mathbb{K}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}} \cdot {\bf D}^{(0)} \;, \\
   \widetilde{\mathbb{L}}  &\equiv \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \;
 \end{align}
\end{subequations}
%
to simplify the notation. Hence,
%
\begin{equation} \label{e:dmu-l-vector}
  \frac{1}{4} 
 \delta {\BM{\upmu}} %({\bf r}_0) 
   =
   \Tr{ 
    \left[ 
         \widetilde{\mathbb{L}} \cdot \delta {\bf C}
    \right] }
   %
   \equiv \sum_{i}^{\rm occ} \sum_\alpha {\bf l}_{i\alpha} \delta C_{\alpha i} \;,
\end{equation}
%
where the Cartesian vector
%
\begin{equation}\label{e:l-vector}
 {\bf l}_{i\alpha} \equiv \left[ \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \right]_{i\alpha} 
      = \left[  {\bf C}^\dagger \cdot \mathbb{M} \cdot \left( {\bf 1} - {\bf D}^{(0)} \right) \right]_{i\alpha}
\end{equation}
%
and $i$ runs over occupied unperturbed MO's.
Note that the intermediate quantity $\widetilde{\mathbb{L}}$ contains the projector onto
the unoccupied MO space at the absence of the external electric field. It means that the
perturbed solution is constructed from the unperturbed virtual orbitals as well.
Although Eq.~\eqref{e:dmu-l-vector}
is exact up to first order in ${\BM\Delta}$, it cannot be inverted at this moment to obtain $\delta {\bf C}$
due to the presence of trace operation, unless the electric field is uniform.
Let the induced dipole moment be partitioned into separate contributions
from the occupied localised MO's (LMO's), i.e.,
%
\begin{equation} \label{e:dmu-part}
 \delta {\BM{\upmu}} = \sum_i^{\rm occ} \delta {\BM{\upmu}}_i \;,
\end{equation}
%
where
%
\begin{equation} \label{e:dmu-dpol}
 \delta {\BM{\upmu}}_i = {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) \;.
\end{equation}
%
In Eq.~\eqref{e:dmu-dpol}, ${\BM{\alpha}}_i$ is the distributed dipole\hyp{}dipole polarisability tensor
associated with the $i$th LMO whereas ${\bf F}({\bf r}_i)$ 
is the electric field evaluated at the $i$th LMO charge centroid,
${\bf r}_i = \tbraket{\phi_i}{\hat{\bf r}}{\phi_i}$.
By inserting the above result into Eq.~\eqref{e:dmu-l-vector},
recasting the right hand side in a matrix form
and assuming real quantities,
one can separate the contributions coming from occupied MO's as
%
\begin{equation} \label{e:dmu-l-vector-mo-transform-X}
 \frac{1}{4} {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) 
   =
   \delta {\bf c}_i^T \cdot {\bf L}_i \;.
\end{equation}
%
Note that now ${\bf L}_i$ is a matrix of dimension $n \times 3$ and $\delta{\bf c}_i$ is the $i$th column of the
change in the LCAO-MO matrix due to the polarisation process.
The solution for $\delta {\bf C}$ is
%
\begin{equation} \label{e:mu-ind.matrix.linear.solution}
  \delta {\bf c}_i^T = \frac{1}{4}
            {\bf F}({\bf r}_i)  \cdot {\BM{\alpha}}_i \cdot 
                    \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \;,
\end{equation}
%
where $\left[ {\bf L}_i  \right]^{-1}_{\rm Left}$ is a left inverse
of ${\bf L}_i$ matrix,
%
\begin{equation} 
      \left[ {\bf L}_i  \right]^{-1}_{\rm Left}   \equiv
       \left[ {\bf L}_i^T \cdot {\bf L}_i \right]^{-1} \cdot {\bf L}_i^T   
\end{equation}
%
with the dimension of $3\times n$. Note that the right inverse of ${\bf L}_i$
does not exist because ${\bf L}_i \cdot {\bf L}_i^T$ is singular due to the fact that usually
$n > 3$.

@@

\begin{acknowledgments}
This project is carried out under POLONEZ programme which has received funding from the European Union's
Horizon~2020 research and innovation programme under the Marie Skłodowska-Curie grant agreement 
No.~665778. \InlineImage{eu-logo.eps} This project is funded by National Science Centre, Poland 
(grant~no. 2016/23/P/ST4/01720) within the POLONEZ 3 fellowship. 
%Many thanks to Robert W.~G{\'o}ra and Robert Zale{\'s}ny 
%for reading and commenting on the manuscript.
\end{acknowledgments}
